Missing data: Current practice in football research and recommendations for improvement

### 2020-09-05 paper/paper.Rmd
change wording in the discussion.
edit wording in the discussion.

Practice to avoid: *Unreported missing values.* While not always the case, even if
not reported, missing data can be obvious. A recent paper examined the
relationship between match performance indicators and outcome, in
Australian Football between the 2001 and 2016. Of the 91 team performance indicators included in the analysis,
at least one variable (i.e., 'meters gained') was not available over the entire 2001-2016 period [@Young2019]**Check this reference**.
This was not noted anywhere in the paper, or in the supplementary
materials. It is unknown whether imputation was undertaken. Without
these details, it is difficult for other analysts and researchers to use
and/or extend the ideas presented in the paper, or reproduce the
analysis. It is also difficult for a reader to evaluate how the
missingness could bias the results. The absence of a
missing data statement when data are missing is not unique to the study
above [@Young2019]**Check this reference**. Authors should include a missing data statement,
irrespective of whether there is data missing, or not.

@@

Practice to avoid: *Not evaluating the effect, or choice, of imputation.* The
imputation method used to impute missing data has the potential to
affect the outcome of a study. Therefore, it is important that authors
understand, and document, how imputation affected the goal of the
analysis (i.e., inference, prediction, or both). For example,
understanding the implications in using mean imputation, compared to
K-nearest neighbour imputation -- the effect on parameter estimates
('significant' versus 'not significant'), and the (un)certainty of the
coefficients (i.e., less/more). When using imputation, authors need to
perform a sensitivity analysis (ref circulation article-Nick). This is
discussed below.

@@

Broadly, there are two types of missing values in data: implicit and
explicit missings. Explicit values are missing, but recorded; whereas
with implicit missinges, their presence is implied based on other
information in the data. For example, in  Table \@ref(tab:implicit-missings), player Koenen has
missing values for quarters two and four. Sometimes values like these
can be logically imputed, as it might be known that these values are
recorded as 0 in this format, rather than NA. Other places missing data
can arise include: bad joins when merging data without corresponding
values, surveys, an inability to collect a biological sample (e.g.,
venous blood), equipment malfunction or failure, or or not being worn.

@@

We recommend that researchers use missing data overview graphics (see
Graphics section in [@Tierney2018], For example, overview plots [@Tierney2017].

@@

To learn more about exploring missing values, we recommend the vignettes
in naniar [@naniar], and the methods in (@Tierney2018)

@@

Imputing data might feel wrong, as we are 'making up data'. The truth
is, we (generally) can never know what the missing values were. The goal
of imputing data is to make the best possible inference from the data.
We recommend data is imputed, with a few caveats. Consider removing
variables with a high proportion of missing data (e.g., variables with
the majority or more missing than not). With respect to a 'rule of
thumb' for when data should be imputed, we deliberately do not suggest
any specific threshold (e.g., impute values when >5% of data are
missing otherwise delete rows with <5% missing) and caution against
such practice. Rather, we recommend values should generally be imputed.

Missing values can occur in the both the predictors (the "independent variables") and response variable (the "dependent variables"). Care should be taken when imputing data, as typically only predictors should be imputed. For discussion on imputing the response value, see ... [@refs]. (note: take a look again)

We recommend avoiding imputation methods that impute the same value,
such as mean or median imputation, and instead suggest using imputation
methods such as: linear regression, k nearest neighbours, or expectation
maximisation.

Much in the way of there is no single 'best' statistical method, there
is no perfect, one-size-fits-all approach for imputing data. The goal is
to generate similar values that might have been otherwise recorded -
sometimes this means using a neighborhood approach of finding similar
values, other times this means predicting responses using linear or tree
models, and other times the most likely value might have been 0 or the
last or first value carried forward. For detailed descriptions and a
summary of these methods, we suggest [@Van_Buuren2012; @Schafer2002; @Cheema2014].

@@

No matter the imputation method used, it is essential to compare results
of different missing data handling to understand how they may bias the
results. For example, comparing analysis results from applying listwise
deletion, compared to mean imputation, compared to linear regression
imputation. This can reveal bias occurring in imputation methods. An
example of this approach is described in the Case Study in
@Tierney2018.

@@

The current state of reporting missing data in the sports literature has room for improvement. In this section we discuss recommended practices for exploring and
reporting missing data. Our intention is not to provide a 'cookbook'
style approach to missing data, but rather broad recommendations to help
researchers when writing the methods and results sections of a study,
and assist researchers when evaluating a study during the peer-review
process.

@@

More often than not, researchers will be faced with missing data. The
reasons for the missingness, along with the decision of how missing data
are treated, can have a profound effect on the validity and accuracy of
a study's results. A survey of 2019 articles on football related topics
showed that current practice is poor. Only about one in 20 studies
reported missing data. To help address this issue, we have discussed
recommendations for exploring and reporting missing data, to assist
researchers during this important process in the analytical pipeline.
Research should consider these recommendations, and pay greater
attention to missing data and its influence on research results.

>>>>

Practice to avoid: *Unreported missing values.* While not always the case, even if
not reported, missing data can be obvious. A recent paper examined the
relationship between match performance indicators and outcome, in
Australian Football between the 2001 and 2016. Of the 91 team performance indicators included in the analysis,
at least one variable (i.e., 'meters gained') was not available over the entire 2001-2016 period [@Young2019]**Check this reference**. **We might need to add a sentence about 'question mark' in the figures?**
This was not noted anywhere in the paper, or in the supplementary
materials. It is unknown whether imputation was undertaken. Without
these details, it is difficult for other analysts and researchers to use
and/or extend the ideas presented in the paper, or reproduce the
analysis. It is also difficult for a reader to evaluate how the
missingness could bias the results. The absence of a
missing data statement when data are missing is not unique to the study
above [@Young2019]**Check this reference**. Authors should include a missing data statement,
irrespective of whether there is data missing, or not.

@@

Practice to avoid: *Not evaluating the effect, or choice, of imputation.* The 
method used to impute missing data has the potential to
affect the outcome of a study. It is important that authors
understand, and document, how imputation affected the goal of the
analysis (i.e., inference, prediction, or both). For example,
understanding the implications in using mean imputation, compared to
K-nearest neighbour imputation--in terms of the effect on parameter estimates
('significant' versus 'not significant'), and the (un)certainty of the
coefficients (i.e., less/more). When using imputation, authors need to
perform a sensitivity analysis (**ref circulation article-Nick**). This is
discussed below.

@@

Broadly, there are two types of missing values in data: implicit and
explicit missings. Explicit values are missing, but recorded; whereas
with implicit missinges, their presence is implied based on other
information in the data. For example, in  Table \@ref(tab:implicit-missings), player Koenen has
missing values for quarters two and four. Sometimes values like these
can be logically imputed, as it might be known that these values are
recorded as 0 in this format, rather than NA. Other places missing data
can arise include: through joins when merging data without corresponding
values, surveys, an inability to collect a biological sample (e.g.,
venous blood), equipment malfunction, failure, or not being worn.

@@

We recommend that researchers use missing data overview graphics (see
Graphics section in [@Tierney2018]; for example, overview plots [@Tierney2017].

@@

To learn more about exploring missing values, we recommend the vignettes
in naniar [@naniar], and the methods in (@Tierney2018).

@@

Imputing data might feel wrong, as we are 'making up data'. The truth
is, we (generally) can never know what the missing values were. The goal
of imputing data is to make the best possible inference from the data.
We recommend data is imputed, with a few caveats. Consider removing
variables with a high proportion of missing data (e.g., variables with
the majority or more missing than not). With respect to a 'rule of
thumb' for when data should be imputed, we deliberately do not suggest
any specific threshold (e.g., impute values when >5% of data are
missing otherwise delete rows with <5% missing) and caution against
such practice. Rather, we recommend values should generally be imputed. 
We recommend avoiding imputation methods that impute the same value,
such as the mean or median, and instead suggest using imputation
methods such as: linear regression, k-nearest neighbours, or expectation
maximisation. Missing values can occur in the both the predictors (the "independent variables") 
and response variable (the "dependent variables"). Care should be taken 
when imputing data, as typically only predictors should be imputed. 
For discussion on imputing the response value, see ... [@refs]. **(note: take a look again)**

Much in the way of there is no single 'best' statistical method, there
is no perfect, one-size-fits-all approach for imputing data. The goal is
to generate similar values that might have been otherwise recorded.
Sometimes this means using a neighborhood approach of finding similar
values. Or it could mean predicting responses using linear or tree
models. Other times the most likely value might have been 0 or the
last or first value carried forward. For detailed descriptions, and a
summary of these methods, we suggest [@Van_Buuren2012; @Schafer2002; @Cheema2014].

@@

Irrespective of the imputation method used, it is essential to compare results
of different missing data handling to understand how they may bias the
results. For example, comparing analysis results from applying listwise
deletion, compared to mean imputation, compared to linear regression
imputation. This can reveal bias occurring in imputation methods. An
example of this approach is described in the Case Study in
@Tierney2018.

@@

The current state of reporting missing data in the sports literature has room for improvement. In this section we discuss recommended practices for exploring and
reporting missing data. Our intention is not to provide a 'cookbook'
style approach to missing data, but rather, broad recommendations to help
researchers when writing the methods and results sections of a study,
and assist researchers when evaluating a study during the peer-review
process.

@@

Our survey of 2019 articles on football related topics
showed that current practice of reporting missing data is poor, 
with only about one in 20 studies reported missing data. To address this issue,
we have provided recommendations for reporting and exploring missing data, to assist researchers 
during this important process in the analytical pipeline.
Research should consider these recommendations, and pay greater
attention to missing data and its influence on research results.
